---
title: "Seurat_to_AnnData"
output: html_document
date: '2022-12-12'
---

Notes:

--Select HVG before making anndata object 
Based on source code, sciPENN filters out cells with low expression & genes with low expression; does median - normalization of genes + protein; does log1p normalization; and then select 1000 HVGs
We could just use gene_list argument in sciPENN_API to set the 1000 HVGs, or we could do all the normalization steps in advance. 
I'm going to do all the normalization steps in Seurat, just to be consistent with the choice of HVGs
(I checked to make sure the Seurat normalization matches the sc.pp normalization results using the same toy data)

--create new variable that is a combo of donor & day
Day is essentially acting as a batch (I think), so I made 9 "batches" - one for each combination of day/donor

--remove control protein data
32,33,34,35, 85, 86 are mouse/rat antibodies
`rownames(prot)[c(32:35, 85:86)]`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
# BiocManager::install("zellkonverter")
library(zellkonverter)
library(SingleCellExperiment)
```

```{r OLD VERSION}
# read in raw protein data
sparse_prot = readRDS('/home/asmauger/biostat625final/sparse_prot.rds')
all.equal(dim(sparse_prot), c(140, 70988))

# read in metadata
metadata = read.csv('/home/asmauger/biostat625final/metadata.csv', header=T)
rownames(metadata) = metadata[,1]
metadata$daydonor = paste0('day',metadata$day, metadata$donor)

# read in raw RNA training data
sparse_RNA_train = readRDS('/home/asmauger/biostat625final/sparse_RNA_train.rds')
# temporary - choose 1000 genes
sparse_RNA_train = sparse_RNA_train[10000:11000,]

# read in raw RNA testing data
sparse_RNA_test = readRDS('/home/asmauger/biostat625final/sparse_RNA_test.rds')
# temporary - choose 1000 genes
sparse_RNA_test = sparse_RNA_test[10000:11000,]

# read in raw protein training data
sparse_prot_train = readRDS('/home/asmauger/biostat625final/sparse_prot_train.rds')

# create Seurat object 
seuratobj = CreateSeuratObject(counts = sparse_prot, meta.data=metadata, assay="prot")
seuratobj2 = CreateSeuratObject(counts= sparse_RNA_train, meta.data=metadata)
seuratobj3 = CreateSeuratObject(counts=sparse_prot_train, meta.data=metadata)
seuratobj4 = CreateSeuratObject(counts= sparse_RNA_test, meta.data=metadata)  
  
# save as single cell experiment
sce <- as.SingleCellExperiment(seuratobj)
sce2 <- as.SingleCellExperiment(seuratobj2)
sce3 <- as.SingleCellExperiment(seuratobj3)
sce4 <- as.SingleCellExperiment(seuratobj4)

# convert to anndata
zellkonverter::writeH5AD(sce, file='/home/asmauger/biostat625final/test.h5ad')
zellkonverter::writeH5AD(sce2, file='/home/asmauger/biostat625final/rna_train_trial.h5ad')
zellkonverter::writeH5AD(sce3, file='/home/asmauger/biostat625final/prot_train_trial.h5ad')
zellkonverter::writeH5AD(sce4, file='/home/asmauger/biostat625final/rna_test_trial.h5ad')
```

```{r NEW VERSION}
# read in metadata
metadata = read.csv('/home/asmauger/biostat625final/metadata.csv', header=T)
rownames(metadata) = metadata[,1]
metadata$daydonor = paste0('day',metadata$day, metadata$donor)

# read in raw RNA training data
sparse_RNA_train = readRDS('/home/asmauger/biostat625final/sparse_RNA_train.rds')

# read in raw RNA testing data
sparse_RNA_test = readRDS('/home/asmauger/biostat625final/sparse_RNA_test.rds')

# read in raw protein training data
sparse_prot_train = readRDS('/home/asmauger/biostat625final/sparse_prot_train.rds')
rownames(sparse_prot_train)[c(32:35, 85:86)] # remove mouse/rat proteins
sparse_prot_train <- sparse_prot_train[-c(32:35, 85:86),]
dim(sparse_prot_train)

# read in raw protein testing data
sparse_prot_test = readRDS('/home/asmauger/biostat625final/sparse_prot_test.rds')
sparse_prot_test <- sparse_prot_test[-c(32:35, 85:86),] # remove mouse/rat proteins
dim(sparse_prot_test)

## Make seurat object and select HVGs
# I am using log1p normalization based on the sciPENN_API code
# there could be better methods to normalize protein

data_list = list(sparse_RNA_train, sparse_RNA_test, sparse_prot_train, sparse_prot_test)
seurat_list = list(0,0,0,0)
for(i in 1:4) {
  seuratobj = CreateSeuratObject(counts = data_list[[i]], meta.data=metadata)
  scale = median(seuratobj$nCount_RNA) # scanpy normalizes to the median
  seuratobj = NormalizeData(seuratobj, normalization.method = "LogNormalize", scale.factor = scale)
  seurat_list[[i]] = seuratobj
}

# I believe sciPENN chooses HVG from a combined dataset, so here I merge the training/testing RNA
all_RNA = merge(x = seurat_list[[1]], y = seurat_list[[2]]) 
                
all_RNA = FindVariableFeatures(all_RNA, nfeatures = 1000) # sciPENN uses 1000
  # the scanpy version does some batch correction, but I don't see an option for this in Seurat

mygenes = VariableFeatures(all_RNA)
reduced_train_data = GetAssayData(seurat_list[[1]])[mygenes,]
reduced_test_data = GetAssayData(seurat_list[[2]])[mygenes,]

## Now convert to anndata (seurat -> SCE -> anndata)

seuratobj = CreateSeuratObject(counts = reduced_train_data, meta.data=metadata)
seuratobj2 = CreateSeuratObject(counts = reduced_test_data, meta.data=metadata)
seuratobj3 = seurat_list[[3]]
seuratobj4 = seurat_list[[4]]

sce <- as.SingleCellExperiment(seuratobj)
sce2 <- as.SingleCellExperiment(seuratobj2)
sce3 <- as.SingleCellExperiment(seuratobj3)
sce4 <- as.SingleCellExperiment(seuratobj4)

zellkonverter::writeH5AD(sce, file='/home/asmauger/biostat625final/rna_train_hvg.h5ad')
zellkonverter::writeH5AD(sce2, file='/home/asmauger/biostat625final/rna_test_hvg.h5ad')
zellkonverter::writeH5AD(sce3, file='/home/asmauger/biostat625final/prot_train.h5ad', X_name='logcounts')
zellkonverter::writeH5AD(sce4, file='/home/asmauger/biostat625final/prot_test.h5ad', X_name='logcounts')

```
```{r scale by batch}

# make scaled (unit variance, 0 mean) versions of data

seuratobj <- ScaleData(seuratobj, split.by='daydonor')
seuratobj2 <- ScaleData(seuratobj2, split.by='daydonor')
seuratobj3 <- ScaleData(seuratobj3, split.by='daydonor')
seuratobj4 <- ScaleData(seuratobj4, split.by='daydonor')

# convert to single cell experiment

sce <- as.SingleCellExperiment(seuratobj)
sce2 <- as.SingleCellExperiment(seuratobj2)
sce3 <- as.SingleCellExperiment(seuratobj3)
sce4 <- as.SingleCellExperiment(seuratobj4)

# convert to anndata

zellkonverter::writeH5AD(sce, file='/home/asmauger/biostat625final/rna_train_hvg_scaled.h5ad', X_name='scaledata')
zellkonverter::writeH5AD(sce2, file='/home/asmauger/biostat625final/rna_test_hvg_scaled.h5ad', X_name = 'scaledata')
zellkonverter::writeH5AD(sce3, file='/home/asmauger/biostat625final/prot_train_scaled.h5ad', X_name='scaledata')
zellkonverter::writeH5AD(sce4, file='/home/asmauger/biostat625final/prot_test_scaled.h5ad', X_name='scaledata')
```

